---
title: "The Impact of Covid-19 on Türkiye"
---

# 1. Project Overview and Scope

The Pandemic has had profound effects on the society and economy worldwide. It can be said that with the pandemic our lives had changed significantly and maybe some can not even remember what the life was like before pandemic. This had positive and negative impacts on the countries. This project investigates such impacts with the scope of Türkiye. The Turkish Statistical Institute (TUIK) databases will be used during this analysis of the Türkiye's situation regarding to the impact of Covid-19 on Türkiye. The main objective is to understand the reasons and structural changes that might have happens after the Covid-19 by comparing the pre and post pandemic periods.

This project will mainly focuses on the population growth, birth and death rates, martial behaviors employment situations, trading activities. The focus will be on 3 periods 2018-2019 the pre-pandemic period, 2020 - 2022 pandemic period and lastly 2023-2024 post-pandemic period. Even though there is multiple years data on the TUIK's database i narrowed the years to these fundamental 7 years to avoid any seasonality of the previous years recession.

# 2. Data

The data sets that are used are from the TUIK. The below is the details of the such structure of these data sets. Also, We need to add a new column to each database since we want the analysis to be focused on 3 time periods.

*-\> Here i will identify the libraries and the structure of the columns to be added later on.*

```{r}

#Getting the libraries needed
library(tidyverse)
library(readxl)
library(lubridate)
library(ggplot2)
library(scales)

#structuring the Columns to include the new periods
add_covid_period <- function(df) {
  df %>% mutate(CovidPeriod = case_when(
    Year %in% 2018:2019 ~ "Pre-COVID",
    Year %in% 2020:2022 ~ "During COVID",
    Year %in% 2023:2024 ~ "Post-COVID",
    TRUE ~ NA_character_
  ))
}
```

### **1- Population Data**

The population data will be used to identify the population trends, detect if there is a slowdown of population growths during or after the Covid-19.

*-\>Here i will import the population data then add the new predefined column and show what the dataset looks like*

```{r}
#| echo: false
population_data <- read_excel("Data/population.xls")
population_data <- add_covid_period(population_data)
head(population_data)

```

## **2- Marriage Statistics**

The Marriage usually used as a symbol of happiness and hope. The increase of marriages usually means that there is a general happy period of a countries. With the trend of the marriages we can understand the welfare of the country, the unemployment rate of the youth and the populations age dsitribution.If the marriages are in a increase trend we can say that the economic conditions of the country is getting better. That's what we will be investşigating with this data set.

*-\>Here i will import the population data then add the new predefined column and show what the dataset looks like*

```{r}
#| echo: false
marriages_data <- read_excel("Data/marriages.xlsx")
marriages_data <- add_covid_period(marriages_data)
head(marriages_data)
```

## **3- Average Age of Marriage**

The average of the marriage along with the total amount of marriages in a country can show us the indebth details of the countries economic conditions. If the average age of marriage is decreasing we might say that the age distribution of the population is moving towards the youths.

*-\>Here i will import the average age of marriage data then add the new predefined column and show what the dataset looks like*

```{r}
#| echo: false
marriages_age_data <- read_excel("Data/marriages_age.xlsx")
marriages_age_data <- add_covid_period(marriages_age_data)
head(marriages_age_data)
```

## **4- Birth Statistics**

The birth statistics are important for a country since increasing the youth of a country can push its growth further. The trends of the birth can means that the country will be increasing the youth population.

*-\>Here i will import the average age of marriage data then add the new predefined column and show what the dataset looks like*

```{r}
#| echo: false
births_data <- read_excel("Data/births.xlsx")
births_data <- add_covid_period(births_data)
head(births_data)
```

## **5- Fertility Rate**

This with the birth statics we can understand the distribution fo the population much better.

*-\>Here i will import the average age of marriage data then add the new predefined column and show what the dataset looks like*

```{r}
#| echo: false
fertility_data <- read_excel("Data/fertility.xlsx")
fertility_data <- add_covid_period(fertility_data)
head(fertility_data)
```

## **6- Death Statistics**

The death is an important factor to understand the affects of the Covid-19 to understand if the rates have been increased or not during or after the pandemic compared to the pre-pandemic era. With this information we can understand our immune system as a population and understand is there is any incline of certain causes of death or are we back to pre-pandemic period.

*-\>Here i will import the population data then add the new predefined column and show what the dataset looks like*

```{r}
#| echo: false
deaths_data <- read_excel("Data/deaths.xlsx")
deaths_data <- add_covid_period(deaths_data)
head(deaths_data)
```

## **7- Employment Data**

This is also one of the important aspect of a countries economical indicators. The higher the employment rate the country's internal investment are distributed. Ideally a country wants to be %100 employed. This data will be used to analyze this situation.

*-\>Here i will import the population data then add the new predefined column and show what the dataset looks like*

```{r}
#| echo: false
employment_data <- read_excel("Data/employment.xlsx")
employment_data <- add_covid_period(employment_data)
head(employment_data)

```

## **8- Export Data**

The exporting of goods and services has a similar indentations of the import. Here the increase in exports means that the country is well off şn the domestic side has sufficient diplomatic communicatin to be able to export its surplus. The trend of the exports also will give us the economical back ground of the Türkiye for the time period.

*-\>Here i will import the population data then add the new predefined column and show what the dataset looks like*

```{r}
#| echo: false
export_data <- read_excel("Data/export.xlsx")
export_data <- add_covid_period(export_data)
head(export_data)
```

## **9- Import Data**

The importing goods and services can mean that the country is not sufficiently producing the goods that is imported. This can be a good sign of diplomacy meanwhile if a country is importing all of its good we can say that this country is economically dependent. With the trend of the imports we can see if Türkiye is getting more depended with its economy.

*-\>Here i will import the population data then add the new predefined column and show what the dataset looks like*

```{r}
#| echo: false
import_data <- read_excel("Data/import.xlsx")
import_data <- add_covid_period(import_data)
head(import_data)
```

## 2.1 Preprocessing

For the preprocessing of the data I used basic excel skills since i create the structure of data i want to analyze. Also the data sets weer small in comparison to other topic i could have analyzed. What i did basically is tidying data via the tidy methodologies of the Hadley Wickham (1). I formatted my data into his ideology of data shaping this way it was much easier to use the data sets i obtained from TUIK.

# 3. Analysis

Once we go through the data and the explanation of the data sets i will use. We can move on with the analysis. The analysis will mainly take 7 sections. These section will separately analyzed then a conclusion will be made for each section.

## 3.1 Population

```{r}
population_summary <- population_data %>%
  group_by(CovidPeriod) %>% 
  summarise(TotalPopulation = sum(Population, na.rm = TRUE), Count = n())
print(population_summary)

```

```{r}

ggplot(population_data, aes(x = CovidPeriod, y = Population, col = CovidPeriod)) +
  geom_point() + scale_y_log10(labels = comma) 
  labs(title = "Population Distribution by COVID Period")



```

```{r}
#| message: false
#| warning: false

ggplot(population_data, aes(x = CovidPeriod, y = Population, col = Gender)) +
  geom_point() + scale_y_log10(labels = comma) +
  facet_wrap(~Gender) +
  labs(title = "Population by Gender and COVID Period")
```

## 3.2 Marriage

```{r}
marriage_summary <- na.omit(marriages_data) %>%
  group_by(CovidPeriod) %>%
  summarise(MeanMarriages = mean(Marriages, na.rm = TRUE), Count = n())
print(marriage_summary)
```

```{r}
ggplot(na.omit(marriages_data), aes(x = CovidPeriod, y = Marriages, fill = CovidPeriod)) +
  geom_boxplot() + scale_y_log10()
  labs(title = "Marriage Count by COVID Period")
```

```{r}
ggplot(na.omit(marriages_age_data), aes(x = Gender, y = Average_Marriage_Age, fill = Gender)) +
  geom_boxplot() +
  facet_wrap(~CovidPeriod) +
  labs(title = "Average Marriage Age by Gender and COVID Period")

```

## 3.4 Birth and Fertility

```{r}
birth_summary <- births_data %>%
  group_by(CovidPeriod) %>%
  summarise(TotalBirths = sum(Births, na.rm = TRUE), Count = n())
print(birth_summary)
```

```{r}
ggplot(births_data, aes(x = CovidPeriod, y = Births, fill = Gender_Baby)) +
  geom_boxplot() + scale_y_log10(labels = comma) +
  facet_wrap(~Gender_Baby) +
  labs(title = "Births by Baby Gender and COVID Period")
```

```{r}
ggplot(na.omit(fertility_data), aes(x = CovidPeriod, y = Fertility_Rate, col = CovidPeriod)) +
  geom_point(size=5) +
  labs(title = "Fertility by COVID Period")
```

## 3.5 Death

```{r}
death_summary <- deaths_data %>%
  group_by(CovidPeriod) %>%
  summarise(TotalDeaths = sum(Deaths_Amount, na.rm = TRUE), Count = n())
print(death_summary)

```

```{r}

ggplot(deaths_data, aes(x = CovidPeriod, y = Deaths_Amount, fill = Gender)) +
  geom_boxplot() + scale_y_log10(labels = comma)+
  facet_wrap(~Gender) +
  labs(title = "Deaths by Gender and COVID Period")


```

```{r}
ggplot(deaths_data, aes(x = Death_Reason, y = Deaths_Amount, col = Death_Reason)) +
  geom_point() +
  facet_wrap(~CovidPeriod) +
  labs(title = "Cause of Deaths by COVID Period") + theme(legend.position = "bottom")
```

## 3.6 Employment

```{r}
employment_summary <- na.omit(employment_data) %>%
  group_by(CovidPeriod) %>%
  summarise(MeanEmployment = mean(Employment_Rate, na.rm = TRUE), Count = n())
print(employment_summary)


```

```{r}
ggplot(na.omit(employment_data), aes(x = CovidPeriod, y = Employment_Rate, col = Gender)) +
  geom_point() +
  facet_wrap(~Status) +
  labs(title = "Employment Rate by Gender, Marital Status and COVID Period")
```

## 3.7 Export and Import

```{r}
export_summary <- na.omit(export_data) %>%
  group_by(CovidPeriod) %>%
  summarise(MeanExport = mean(Export, na.rm = TRUE), Count = n())
print(export_summary)


```

```{r}
import_summary <- na.omit(import_data) %>%
  group_by(CovidPeriod) %>%
  summarise(MeanImport = mean(Import, na.rm = TRUE), Count = n())
print(import_summary)

```

```{r}

ggplot(na.omit(export_data), aes(x = CovidPeriod, y = Export, fill = CovidPeriod)) +
  geom_boxplot() + scale_y_log10(labels = comma) +
  labs(title = "Export Amount by COVID Period")


```

```{r}
ggplot(na.omit(import_data), aes(x = CovidPeriod, y = Import, fill = CovidPeriod)) +
  geom_boxplot() + scale_y_log10 (labels = comma) +
  labs(title = "Import Amount by COVID Period")
```

# 4. Results and Key Takeaways
