---
title: "The Impact of Covid-19 on Türkiye"
---

# 1. Project Overview and Scope

The Pandemic has had profound effects on the society and economy worldwide. It can be said that with the pandemic our lives had changed significantly and maybe some can not even remember what the life was like before pandemic. This had positive and negative impacts on the countries. This project investigates such impacts with the scope of Türkiye. The Turkish Statistical Institute (TUIK) databases will be used during this analysis of the Türkiye's situation regarding to the impact of Covid-19 on Türkiye. The main objective is to understand the reasons and structural changes that might have happens after the Covid-19 by comparing the pre and post pandemic periods.

This project will mainly focuses on the population growth, birth and death rates, martial behaviors employment situations, trading activities. The focus will be on 3 periods 2018-2019 the pre-pandemic period, 2020 - 2022 pandemic period and lastly 2023-2024 post-pandemic period. Even though there is multiple years data on the TUIK's database i narrowed the years to these fundamental 7 years to avoid any seasonality of the previous years recession.

# 2. Data

The data sets that are used are taken from the TUIK. The below is the details of the such structure of these data sets.

```{r}
#| include: false
#Getting the libraries needed
library(tidyverse)
library(readxl)
library(lubridate)
library(ggplot2)
```

**1- Population Data**

The population data will be used to identify the population trends, detect if there is a slowdown of population growths during or after the Covid-19.

```{r}
#| echo: false
population_data <- read_excel("Data/population.xls")
head(population_data)

```

**2- Death Statistics**

The death is an important factor to understand the affects of the Covid-19 to understand if the rates have been increased or not during or after the pandemic compared to the pre-pandemic era. With this information we can understand our immune system as a population and understand is there is any incline of certain causes of death or are we back to pre-pandemic period.

```{r}
#| echo: false
deaths_data <- read_excel("Data/deaths.xlsx")
head(deaths_data)
```

**3- Marriage Statistics**

The Marriage usually used as a symbol of happiness and hope. The increase of marriages usually means that there is a general happy period of a countries. With the trend of the marriages we can understand the welfare of the country, the unemployment rate of the youth and the populations age dsitribution.If the marriages are in a increase trend we can say that the economic conditions of the country is getting better. That's what we will be investşigating with this data set.

```{r}
#| echo: false
marriages_data <- read_excel("Data/marriages.xlsx")
head(marriages_data)
```

**4- Average Age of Marriage**

The average of the marriage along with the total amount of marriages in a country can show us the indebth details of the countries economic conditions. If the average age of marriage is decreasing we might say that the age distribution of the population is moving towards the youths.

```{r}
#| echo: false
marriages_age_data <- read_excel("Data/marriages_age.xlsx")
head(marriages_age_data)
```

**5- Import Data**

The importing goods and services can mean that the country is not sufficiently producing the goods that is imported. This can be a good sign of diplomacy meanwhile if a country is importing all of its good we can say that this country is economically dependent. With the trend of the imports we can see if Türkiye is getting more depended with its economy.

```{r}
#| echo: false
import_data <- read_excel("Data/import.xlsx")
head(import_data)
```

**6- Export Data**

The exporting of goods and services has a similar indentations of the import. Here the increase in exports means that the country is well off şn the domestic side has sufficient diplomatic communicatin to be able to export its surplus. The trend of the exports also will give us the economical back ground of the Türkiye for the time period.

```{r}
#| echo: false
export_data <- read_excel("Data/export.xlsx")
head(export_data)
```

**7- Birth Statistics**

The birth statistics are important for a country since increasing the youth of a country can push its growth further. The trends of the birth can means that the country will be increasing the youth population.

```{r}
#| echo: false
births_data <- read_excel("Data/births.xlsx")
head(births_data)
```

**8- Fertility Rate**

This with the birth statics we can understand the distribution fo the population much better.

```{r}
#| echo: false
fertility_data <- read_excel("Data/fertility.xlsx")
head(fertility_data)
```

**9- Employment Data**

This is also one of the important aspect of a countries economical indicators. The higher the employment rate the country's internal investment are distributed. Ideally a country wants to be %100 employed. This data will be used to analyze this situation.

```{r}
#| echo: false

employment_data <- read_excel("Data/employment.xlsx")
head(employment_data)
```

**10- Education Level**

Education level is related with the technological advancements of a country to analyze the root of its economy. Some countries can be tech driven while others are related with the service industries. Checking the trends of the educational level of the country can give us some idea of its economical drivers. Checking the periods education level will also give us clues on the changes in the Türkiye.

```{r}
#| echo: false
education_data <- read_excel("Data/education.xlsx")
head(education_data)
```

## 2.1 Preprocessing

For the preprocessing of the data I used basic excel skills since i create the structure of data i want to analyze. Also the data sets weer small in comparison to other topic i could have analyzed. What i did basically is tidying data via the tidy methodologies of the Hadley Wickham (1). I formatted my data into his ideology of data shaping this way it was much easier to use the data sets i obtained from TUIK.

# 3. Analysis

Once we go through the data and the explanation of the data sets i will use. We can move on with the analysis. The analysis will mainly take 7 sections. These section will seperately analyzed then a conclusion will be made for each section.

## 3.1 Population

The first step of every analysis is to see the general view of the trend. Here we can see the increase of the population over time. With this view we can observe;

-   The overall population is increasing over the time period.

-   The increase of the pre-pandemic and during pandemic periods seems to be similar.

-   There is a drop of the incline of the population growth during the post-pandemic period.

```{r}

population_summary <- population_data %>%
  group_by(Year) %>%
  summarise(TotalPopulation = sum(Population))

ggplot(population_summary, aes(x = Year, y = TotalPopulation)) +
  geom_line(color = "red",linewidth = 1.5) +
  scale_x_continuous(breaks = unique(population_summary$Year)) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Total Population Over Years", y = "Population", x = "Year")


```

We need to dive in more with the population data. Lets see the population with the genders. Here we can see the males as blue and females as pink. the main observations from this graph is;

-   The growth trends of the male and female population is almost the same (there is no significant difference)

-   The male population is much higher than the female population. The possible reason for this situation will be analyzed on the next sections.

-   Even though the male population is above the female population we can see that the difference between the two genders is getting smaller each year. The possible causes for such a think can we answered once we analyzed the birth and the death data sets.

```{r}
#| message: false
#| warning: false
gender_summary <- population_data %>% group_by(Year, Gender) %>% summarise(Population = sum(Population))

ggplot(gender_summary, aes(x = Year, y = Population, color = Gender)) +
  geom_line(size = 1.5) +
  scale_color_manual(values = c("Erkek" = "#1f77b4", "Kadın" = "#e377c2")) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Population Trend by Gender", x = "Year", y = "Population")

```

Now lets analyze the population by province. Here we can see that the top %20 of the TR is giving the population differences. Lets check the top %20 much in depth.

```{r}

top10_provinces <- population_data %>%
  group_by(City) %>%
  summarise(AveragePopulation = mean(Population)) %>%
  arrange(desc(AveragePopulation)) %>%
  slice_head(n = 81)

ggplot(top10_provinces, aes(x = reorder(City, -AveragePopulation), y = AveragePopulation, fill = City)) +geom_col(show.legend = FALSE) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Cities by Average Population (2018–2024)", x = "City", y = "Average Population") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

The graph 2 shows a better details. I can say that the top 5 cities would be showing the most of the TRs affects. (Pareto principle) We can now go in depth with the top 5 cities over the year trends.

```{r}
top10_provinces <- population_data %>%
  group_by(City) %>%
  summarise(AveragePopulation = mean(Population)) %>%
  arrange(desc(AveragePopulation)) %>%
  slice_head(n = 16)

ggplot(top10_provinces, aes(x = reorder(City, -AveragePopulation), y = AveragePopulation, fill = City)) +geom_col(show.legend = FALSE) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Top 20% Provinces by Average Population (2018–2024)", x = "City", y = "Average Population") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Except for İstanbul we can say that the population is almost increasing linearly. I see that the yearly trends of the Türkiye is depended on the Istanbul Population.

```{r}
# Identify top 5 provinces by total population over all years
top5_provinces <- population_data %>%
  group_by(City) %>%
  summarise(TotalPopulation = sum(Population)) %>%
  arrange(desc(TotalPopulation)) %>%
  slice_head(n = 5) %>%
  pull(City)

# Filter the data for these top 5 provinces
top5_trend_data <- population_data %>%
  filter(City %in% top5_provinces) %>%
  group_by(Year, City) %>%
  summarise(Population = sum(Population), .groups = "drop")

# Plot using facet_wrap by Province
ggplot(top5_trend_data, aes(x = Year, y = Population)) +
  geom_line(color = "#1f77b4", size = 1.5) +
  facet_wrap(~ City, scales = "free_y") +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Yearly Population Trend of Top 5 Provinces", x = "Year", y = "Population") +
  theme_minimal()
```

Finally the Gender factor of the each City. We can see much different results compared to the Türkiye View.

```{r}
# Identify top 5 provinces by total population over all years
top5_provinces <- population_data %>%
  group_by(City) %>%
  summarise(TotalPopulation = sum(Population)) %>%
  arrange(desc(TotalPopulation)) %>%
  slice_head(n = 5) %>%
  pull(City)

# Filter and prepare the data for these provinces, grouped by year and gender
top5_gender_trend <- population_data %>%
  filter(City %in% top5_provinces) %>%
  group_by(Year, City, Gender) %>%
  summarise(Population = sum(Population), .groups = "drop")

# Plot using facet_wrap by Province, with line for each gender
ggplot(top5_gender_trend, aes(x = Year, y = Population, color = Gender)) +
  geom_line(size = 1.2) +
  facet_wrap(~ City, scales = "free_y") +
  scale_color_manual(values = c("Erkek" = "#1f77b4", "Kadın" = "#e377c2")) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Yearly Population Trend by Gender in Top 5 Provinces",
       x = "Year", y = "Population", color = "Gender") +
  theme_minimal()
```

## 3.2 Education

xxxxxx

## 3.3 Marriage

xxxxxx

## 3.4 Birth and Fertility

## 3.5 Death

## 3.6 Employment

## 3.7 Export and Import

xxxxxx

# 4. Results and Key Takeaways

xxxxxx
