---
title: "Assignment 1"
---

# Assignment 1

```         
```

My first assignment has three parts.

## (a) Brief Summary of Baykal

He is really friendly and shared his own thoughts about this type of engagements with students and how he would wanted some to introduce what his job life looks like.

In the first section he talked about his experiences after the graduation. Eve though he graduated from Industrial engineering he started to work in the software industry. In his current job he is working in the sales of the developed OR system but he is more technical side of this sales business.

In the second part he discussed what is OR is in the modern technology. The general industry is now focuses more on the AI and Machine Learning rather than the OR itself. Now we can more see the analytics instead of the OR/optimization. He mentioned that the analytics can be observed on 4 main fields. The most basic one is the descriptive analytics which just shows what the data is, second is diagnostic analytics which analyses the root causes of the problems, predictive analytics focuses on the future possibilities and lastly the prescriptive analytics focuses on the action of the resolving the root causes. He also mentioned the tools that can be used for each analytic analyzes. These analytic method can be used together (also this is encouraged). Using all these method together will help to analyze problem, find the root cause and find a solution.

Lastly he mentioned how we can project-ify these methods and how to understand the KPI's. He also encouraged us the deploy the projects we have. This will show what is missing with the project what can be done to improve. The user is the king approach will help with our final outcome.

## (b) A Basic Computation Using "mtcars"

```{r}
compute_stats <- function(x){
  list(Mean = mean(x), Median = median(x), Variance=var(x), InterQuartileRange=IQR(x), Minimum = min(x), 
            Maximum = max(x))
}
head(sapply(mtcars,compute_stats),100)
```

## (c) Having Fun with "NA_Example"

```{r}
library(dslabs)
data("na_example")
na_example
```

```{r}
TotalNA <- sum(is.na(na_example))
IndexNA <- which(is.na(na_example))
strNA <- paste0('NA ','Amount ','-> ',TotalNA,'\n')
print(strNA)
print('Indeces of the NAs -> ')
print(IndexNA)
```

```{r}
mean <- mean(na_example,na.rm = TRUE)
stdev <- sd(na_example, na.rm=TRUE)
meanStr <- paste0('Mean (w/o NA) ->',mean)
stdevStr <- paste0('Standart deviation (w/o NA) ->',stdev)
print(meanStr)
print(stdevStr)
```

```{r}
mean <- mean(na_example,na.rm = TRUE)
arr1 <- na_example
arr1[is.na(na_example)] <- mean
arr1Str <- paste0('New Dataset With Mean Replacing NAs')
print(arr1Str)
print(arr1)
stdev <- sd(na_example, na.rm=TRUE)
arr2 <- na_example
arr2[is.na(na_example)] <- stdev
arr2Str <- paste0('New Dataset With Standart Deviation Replacing NAs')
print(arr2Str)
print(arr2)

```

```{r}
mean1 <- mean(arr1)
mean2 <- mean(arr2)
std1 <- sd(arr1)
std2 <- sd(arr2)
print('Mean of Mean Replacing NAs')
print(mean1)
print('Mean of StDev Replacing NAs')
print(mean2)
print('StDev of Mean Replacing NAs')
print(std1)
print('StDev of StDev Replacing NAs')
print(std2)
```

[**Comparison**]{.underline}

The mean of the original data-set is 2.30 meanwhile the standard deviation is 1.22.

The mean of the replacement of the NA's with mean value (2.30) is 2.30 and the standard deviation is 1.13.

The mean of the replacement of the NA's with standard devaition value (1.22) is 2.14 and the standard deviation is 1.19.

When we do the comparison i can say that the changing the NAs with a mean results in a same mean value as the original set but different standard deviations. This means that the center of the both data sets are the same (2.30) but the distance of these points to the center is different. In this case the points are located much closer to the center compared to the original set. Similarly when we compare the other set with the original we can say that the center point is different than the original set but the dispersion of the points are much closer to the original set.

[**Decision**]{.underline}

I believe handling the missing data is more depends on what we are trying to achieve with our analysis. If we would like to see what central point is a better solution to our problem.For example we would like to place an access point to the neighborhoods of Ankara, we would choose to focus on the central location of the orders. For this problem i could replace the NAs with the mean. If the solution of the problem involves the dispersion of the points. For example if we would like to control the fragment damage of the milling machine, i would much focus on the dispersion of the fragments instead of their centers. For this problem i would rather choose replacing the NAs with the standard deviation.
